A Bidirected String Graph Model of Genome Assembly

### 2012-11-28 presentation/presentation.tex
presentation

\begin{frame}{Genomes}
	\begin{itemize}
		\item A genome (for our purposes) is a long string of A's, C's, G's, and T's
		\item Genomes are dual-stranded
		\begin{itemize}
			\item {\tt A} always pairs with {\tt T} and {\tt C} always pairs
			with {\tt G}
			\item The two strands run in opposite directions
		\end{itemize}
	\end{itemize}
	\begin{center}
	{\tt $\leftarrow$ GTTGCTATAAATACGGGTCATGGTATTTTAGTACCG $\leftarrow$ \\}
	{\tt $\rightarrow$ CAACGATATTTATGCCCAGTACCATAAAATCATGGC $\rightarrow$}
	\end{center}
	\begin{itemize}
		\item Each letter in the genome is called {\bf base pair} \\(abbreviation:
		{\bf bp})
	\end{itemize}
	\begin{table}[H]
		\caption{Approximate genome sizes of selected organisms}
		\begin{tabular}{|c|c|c|}
			\hline
			{\it E. coli} (a bacterium) &  ~4,600,000 bp \\
			\hline
			{\it Ananas comosus} (pineapple plant) &	~500,000,000 bp \\
			\hline
			{\it Homo sapiens} (human)       &  ~3,000,000,000 bp \\
			\hline
		\end{tabular}
	\end{table}
\end{frame}

@@

\begin{frame}{Reads}
	\begin{itemize}
		\item A {\bf read} is a short sequence of base pairs that represents
		some DNA that was sequenced by the sequencing machine.
		\item A read may come from any location on the genome, from either
		strand.
		\item Reads from opposite strands go in opposite directions.
			\begin{itemize}
				\item As a result, a sequence and its {\bf reverse-complement}
				(the corresponding sequence on the other strand, going
				backwards) are usually considered equivalent.
			\end{itemize}
	\end{itemize}
		{\scriptsize
			{\tt 1.}
			{\tt \color{red} $\leftarrow$TGAGCTTAAGGCTTATCTATCTTCAGACGACA$\leftarrow$ }\\
			{\tt 2.}
			\hspace*{1.76cm}{\tt \color{red} $\leftarrow$TTATCTATCTTCAGACGACTATTATAGCGCGG$\leftarrow$}\\
			{\tt G.}
			{\tt $\leftarrow$TGAGCTTAAGGCTTATCTATCTTCAGACGACTATTATAGCGCGGCCAAGACTACGCGGAGCCCC$\leftarrow$} \\
			{\tt G.}
			{\tt $\rightarrow$ACTCGAATTCCGAATAGATAGAAGTCTGCTGATAATATCGCGCCGGTTCTGATGCGCCTCGGGG$\rightarrow$} \\
			{\tt 3.} \hspace*{3.55cm}{\tt \color{blue} $\rightarrow$TCTGCTGATAATATCGCGCCGGTTCTGATGCG$\rightarrow$}
			\vspace{0.5cm}
		}
		\\
		{\footnotesize Above: a very short genome (G.) and three reads that came from it (1.,
		2., 3.).}
\end{frame}

@@

\begin{frame}{Overlaps between reads}
	Consider two reads $f$ and $g$, and their reverse complements $f'$ and $g'$.
	The reads $f$ and $g$ are said to {\bf exactly overlap} by some length
	$\ell$ if and only if at least one of the following is true:
	\begin{itemize}
		\item The last $\ell$ bp of $f$ exactly match the first $\ell$ bp of
		$g$.
		\item The last $\ell$ bp of $f$ exactly match the first $\ell$ bp of
		$g'$.
		\item The last $\ell$ bp of $f'$ exactly match the first $\ell$ bp of
		$g$.
		\item The last $\ell$ bp of $f'$ exactly match the first $\ell$ bp of
		$g'$.
	\end{itemize}
	An exact overlap $o$ is fully described by the tuple $(f, g, f_{beg},
	g_{beg}, len, rc)$.
	%there $f$ and $g$ are the reads of the overlap,
	%$f_{beg}$ and $g_{beg}$ and the starting position of the overlap on each
	%read, $len$ is the length of the overlap, and $rc$ indicates whether the
	%overlap is reverse-complement or not.
\end{frame}

@@

\begin{frame}{Walking through the bidirected string graph}
	\begin{itemize}
		\item A {\bf walk on a bidirected graph} is a continuous sequence of
		edges such that if any vertex $v$ is entered through a head inwards, it
		is left on a head outwards (unless it is the end of the path), and vice
		versa.

		\item A bidirected edge may be traversed in opposite directions during
		the same walk.

		\item The reverse of a walk in a directed graph is also a bidirected walk.

		\item Interpretation: one direction of the walk spells out the {\bf
		forward sequence}.  The other direction spells out the {\bf reverse
		complement sequence.}
	\end{itemize}
\end{frame}

@@

\begin{frame}{Transitive reduction}
	\begin{itemize}
		\item Very commonly, given three adjacent reads $f$, $g$, and $h$, $f$ will overlap
		$h$ as well as $g$.  This is redundant information because we can equivalently
		walk $f$ to $g$ to $h$ (or $h$ to $g$ to $f$, if going the other way).

@@

\begin{frame}{Transitive reduction (cont.)}
		Transitive reduction algorithm:  Go through each vertex and examine
		neighbors up to 2 steps away to identify all transitive edges leaving this
		vertex.
	\begin{algorithm}[H]
		\caption{{\sc TransitiveReduction}($G$)}
	{\footnotesize
		\begin{algorithmic}
			\FOR{{\bf each} vertex $v \in G$}
				\STATE{Mark all neighbors of $v$ as {\tt INPLAY}.}
				\STATE{Set {\tt longest} $\gets$ length of the longest edge leaving $v$}
				\FOR{{\bf each} edge $v \to w$ in order of length}
					\FOR{{\bf each} edge $w \to x$ in order of length}
						\IF{$x$ is not marked {\tt INPLAY}}
							\STATE{\bf continue}
						\ENDIF
						\IF{$len(v \to w) + len(w \to x) \ge$ {\tt longest}}
							\STATE{\bf break}
						\ENDIF
						\STATE{Mark $x$ as {\tt ELIMINATED}}
					\ENDFOR
				\ENDFOR
				\STATE{Eliminate all edges $v \to w$ where $w$ is marked {\tt ELIMINATED}}
			\ENDFOR
			\STATE{Eliminate all edges marked {\tt ELIMINATED}}
		\end{algorithmic}
	}
	\end{algorithm}
\end{frame}

@@

\begin{frame}{Current state of the graph}
	\begin{itemize}
		\item Non-repetitive genomes and/or long reads: The assembly probably is
		already complete!
	\end{itemize}

@@

\begin{frame}{Current state of the graph (cont.)}
	\begin{itemize}
		\item Repetitive genomes and/or short reads: The graph will still be
		very complicated.
		\item Swap out that 4,600,000 with {\it E. coli}'s genome, which
		contains some repeats, and we get a complicated graph, {\bf even after
		increasing the read length to 1000bp}:
	\end{itemize}
	\begin{center}
		\includegraphics[width=0.6\textwidth]{E_coli.pdf}
	\end{center}
\end{frame}

\begin{frame}{Next steps}
	\begin{itemize}
		\item Given a bidirected string graph that may, in general, have
		thousands of edges and vertices even after transitive reduction and
		collapsing unbranched paths, how can we find the most likely assembly?
		\item Solution: Identify edges that may need to be traversed more than
		once, then solve a minimum-cost network flow problem to find the minimum
		number of edges that need to be traversed multiple times in order to
		find
		\item This is also similar to the Chinese Postman Problem, but this is a
		{\it bidirected graph}, and we are also okay with the final
		reconstruction of the genome still being discontinuous, rather than a
		single path.
	\end{itemize}
\end{frame}

>>>>

\begin{frame}{Genomes}
	\begin{itemize}
		\item A genome (for our purposes) is a long string of A's, C's, G's, and T's
		\item Genomes are dual-stranded
		\begin{itemize}
			\item {\tt A} always pairs with {\tt T} and {\tt C} always pairs
			with {\tt G}
			\item The two strands run in opposite directions
		\end{itemize}
	\end{itemize}
	\begin{center}
	{\tt $\leftarrow$ GTTGCTATAAATACGGGTCATGGTATTTTAGTACCG $\leftarrow$ \\}
	{\tt $\rightarrow$ CAACGATATTTATGCCCAGTACCATAAAATCATGGC $\rightarrow$}
	\end{center}
	\begin{itemize}
		\item Each letter in the genome is called a {\bf base pair} \\(abbreviation:
		{\bf bp})
	\end{itemize}
	\begin{table}[H]
		\caption{Approximate genome sizes of selected organisms}
		\begin{tabular}{|c|c|c|}
			\hline
			{\it E. coli} (a bacterium) &  ~4,600,000 bp \\
			\hline
			{\it Ananas comosus} (pineapple plant) &	~500,000,000 bp \\
			\hline
			{\it Homo sapiens} (human)       &  ~3,000,000,000 bp \\
			\hline
		\end{tabular}
	\end{table}
\end{frame}

@@

\begin{frame}{Reads}
	\begin{itemize}
		\item A {\bf read} is a short sequence of base pairs that represents
		some DNA that was sequenced by the sequencing machine.
		\item A read may come from any location on the genome, from either
		strand.
		\item Reads from opposite strands go in opposite directions.
			\begin{itemize}
				\item As a result, a sequence and its {\bf reverse-complement}
				(the corresponding sequence on the other strand, going
				backwards) are usually considered equivalent.
			\end{itemize}
	\end{itemize}
		{\scriptsize
			{\tt 1.}
			\hspace*{-0.03cm}{\tt\color{red}$\leftarrow$TGAGCTTAAGGCTTATCTATCTTCAGACGACA$\leftarrow$ }\\
			{\tt 2.}
			\hspace*{1.76cm}{\tt \color{red} $\leftarrow$TTATCTATCTTCAGACGACTATTATAGCGCGG$\leftarrow$}\\
			{\tt G.}
			{\tt $\leftarrow$TGAGCTTAAGGCTTATCTATCTTCAGACGACTATTATAGCGCGGCCAAGACTACGCGGAGCCCC$\leftarrow$} \\
			{\tt G.}
			{\tt $\rightarrow$ACTCGAATTCCGAATAGATAGAAGTCTGCTGATAATATCGCGCCGGTTCTGATGCGCCTCGGGG$\rightarrow$} \\
			{\tt 3.} \hspace*{3.55cm}{\tt \color{blue} $\rightarrow$TCTGCTGATAATATCGCGCCGGTTCTGATGCG$\rightarrow$}
			\vspace{0.5cm}
		}
		\\
		{\footnotesize Above: a very short genome (G.) and three reads that came from it (1.,
		2., 3.).}
\end{frame}

@@

\begin{frame}{Overlaps between reads}
	Consider two reads, $f$ and $g$, and their reverse complements $f'$ and $g'$.
	The reads $f$ and $g$ are said to {\bf exactly overlap} by some length
	$\ell$ iff at least one of the following is true:
	\begin{itemize}
		\item The last $\ell$ bp of $f$ exactly match the first $\ell$ bp of
		$g$.
		\item The last $\ell$ bp of $f$ exactly match the first $\ell$ bp of
		$g'$.
		\item The last $\ell$ bp of $f'$ exactly match the first $\ell$ bp of
		$g$.
		\item The last $\ell$ bp of $f'$ exactly match the first $\ell$ bp of
		$g'$.
	\end{itemize}
	An exact overlap $o$ is fully described by the tuple $(f, g, f_{beg},
	g_{beg}, len, rc)$.
	%there $f$ and $g$ are the reads of the overlap,
	%$f_{beg}$ and $g_{beg}$ and the starting position of the overlap on each
	%read, $len$ is the length of the overlap, and $rc$ indicates whether the
	%overlap is reverse-complement or not.
\end{frame}

@@

\begin{frame}{Walks on a bidirected graph}
	\begin{itemize}
		\item A {\bf walk on a bidirected graph} is a continuous sequence of
		edges such that if any vertex $v$ is entered through a head inwards, it
		is left on a head outwards (unless it is the end of the path), and vice
		versa.
		\item A bidirected edge may be traversed in opposite directions during
		the same walk.
		\item The reverse of a walk in a directed graph is also a bidirected walk.
		\item $1 \to 2 \to 3 \to 4$ and $4 \to 3 \to 2 \to 1$ are both valid
		walks in the bidirected graph below.
	\end{itemize}
		\begin{center} {\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
			\node (1) at (0.00, 0) {1};
			\node (2) at (1.75, 0) {2};
			\node (3) at (3.50, 0) {3};
			\node (4) at (5.25, 0) {4};
			\draw[<->,very thick] (1) edge (2);
			\draw[>-<,very thick] (2) edge (3);
			\draw[<-<,very thick] (3) edge (4);
		\end{tikzpicture} } \end{center}
\end{frame}

\begin{frame}{Walking through the bidirected string graph}
	\begin{itemize}
		\item A walk through a bidirected string graph constructed from read
		overlaps models a way in which the reads can be assembled together
		consistently.

		\item Each read may be used in forward or reverse-complement
		orientation, depending on the orientation of the arrow head when the
		corresponding vertex is entered.
	\end{itemize}
	\begin{figure}[H]
		{\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
			\node (1) at (0.00, 0) {1};
			\node (2) at (1.75, 0) {2};
			\node (3) at (3.50, 0) {3};
			\node (4) at (5.25, 0) {4};
			\draw[<->,very thick] (1) edge (2);
			\draw[>-<,very thick] (2) edge (3);
			\draw[<-<,very thick] (3) edge (4);
		\end{tikzpicture} }

		\vspace{2.5mm}

		\begin{tikzpicture}
				\draw[<-,style=very thick] (0.0, 0.0) -- (3.0, 0.0);
				\draw[<-,style=very thick] (0.8, -0.3) -- (3.8, -0.3);
				\draw[->,style=very thick] (1.6, -0.6) -- (4.6, -0.6);
				\draw[<-,style=very thick] (2.4, -0.9) -- (5.4, -0.9);
				\draw (-0.3, 0.0) node {1.};
				\draw (-0.3, -0.3) node {2.};
				\draw (-0.3, -0.6) node {3.};
				\draw (-0.3, -0.9) node {4.};
		\end{tikzpicture}
		\caption{A bidirected graph and the implied assembly of the
		corresponding reads}
	\end{figure}
\end{frame}

@@

\begin{frame}{Transitive reduction}
	\begin{itemize}
		\item Very commonly, given three adjacent reads $f$, $g$, and $h$, $f$
		will overlap $h$ as well as $g$.  This is redundant because we can,
		equivalently, walk $f$ to $g$ to $h$ rather than directly walking from
		$f$ to $h$.

@@

\begin{frame}{Transitive reduction (cont.)}
		Transitive reduction algorithm:  Go through each vertex and examine
		neighbors up to 2 steps away to identify all transitive edges leaving this
		vertex.
	\begin{algorithm}[H]
		\caption{{\sc TransitiveReduction}($G$)}
	{\footnotesize
		\begin{algorithmic}
			\FOR{{\bf each} vertex $v \in G$}
				\STATE{Mark all neighbors of $v$ as {\tt INPLAY}.}
				\STATE{Set {\tt longest} $\gets$ length of the longest edge leaving $v$}
				\FOR{{\bf each} edge $v \to w$ in order of length}
					\FOR{{\bf each} edge $w \to x$ in order of length}
						\IF{$x$ is not marked {\tt INPLAY}}
							\STATE{\bf continue}
						\ENDIF
						\IF{$len(v \to w) + len(w \to x) \ge$ {\tt longest}}
							\STATE{\bf break}
						\ENDIF
						\STATE{Mark $x$ as {\tt ELIMINATED}}
					\ENDFOR
				\ENDFOR
				\STATE{Mark as transitive all edges $v \to x$ where $x$ is marked {\tt ELIMINATED}}
			\ENDFOR
			\STATE{Eliminate all edges marked transitive.}
		\end{algorithmic}
	}
	\end{algorithm}
\end{frame}

@@

\begin{frame}{Current state of the graph}
	\begin{itemize}
		\item Non-repetitive genomes and/or long reads: The assembly is already
		complete!
	\end{itemize}

@@

\begin{frame}{Current state of the graph (cont.)}
	\begin{itemize}
		\item Repetitive genomes and/or short reads: The graph will still be
		very complicated.
		\item Swap out the 4.6 Mbp random genome with {\it E. coli}'s genome,
		which is the same size but contains some repeats, and we get a
		complicated graph, {\bf even after increasing the read length to
		1000bp}:
	\end{itemize}
	\begin{center}
		\includegraphics[width=0.6\textwidth]{E_coli.pdf}
	\end{center}
\end{frame}

\begin{frame}{Next steps}
	\begin{itemize}
		\item Given a bidirected string graph that may, in general, have
		thousands of edges and vertices even after transitive reduction and
		collapsing unbranched paths, how can we find the most likely assembly?
		\item Solution: Identify edges that may need to be traversed more than
		once, then solve a minimum-cost network flow problem to find the minimum
		number of edges that need to be traversed multiple times in order to
		make an Eulerian path.
		\item This is also similar to the Chinese Postman Problem, but this is a
		{\it bidirected graph}, and we are also okay with the final
		reconstruction of the genome still being discontinuous, rather than a
		single path (if necessary).
	\end{itemize}
\end{frame}
